# Comparison across other training data

I want to make sure that the selected variables are not changing a lot with different random test splits.  I'll make 20 random splits and look for variables that seem problematic (effect changes depending on the random data). I am going to select 50% of my data for training instead of 80%. I want to see how the variable selection and importance changes with different training data.

## Set-up

This chapter will use the following libraries.
```{r}
library(biomod2)
library(MASS)
library(ggplot2)
library(ecospat)
library(sp)
library(raster)
```

### Load the data

Load the shapefiles, Trillium data and variables raster stack created earlier.
```{r}
# The VT, NH and HB outlines
load("data/shapefiles.RData")

# The original observation data lat/lon
load("data/trillium_presences.RData")

# All the data needed to fit the SDM
# The grid cells with presences and background grid cells w predictors
# The variables that I'll use for various models
load("data/sdm_data.RData")
# Raster of all predictors
allVars <- raster::brick("data/allVars.grd")
```

## Run fits


```{r}
modlist <- list()
n <- length(traindatlist)
for(i in 1:n){
gamEnv <- gam::gam(formula = gamfm(envvars, 4),
            data=traindatlist[[i]]$traindat, 
            family="binomial")
modlist[[paste("split",i)]] <- gamEnv
}
```

```{r}
df <- data.frame()
for(i in 1:n){
  mod1 <- modlist[[i]]
rp <- biomod2::response.plot2(models = "mod1",
                     Data = datlist[[i]]$traindat,
                     show.variables = envvars,
                     fixed.var.metric = 'mean', plot = FALSE, use.formal.names = TRUE)
df <- rbind(df, cbind(rp, i=paste("samp", i)))
}
```

```{r}
gg.rp <- ggplot(df, aes(x = expl.val, y = pred.val)) +
  geom_line(aes(col=i)) + ylab("prob of occ") + xlab("") + 
  rp.gg.theme + 
  facet_wrap(~ expl.name, scales = 'free_x') +
  ggtitle("Response curves across random samples") +
  theme(legend.position = "none")
print(gg.rp)
```

```{r}
df2 <- data.frame()
for(i in 1:n){
  gamStart <- gam::gam(formula = gamfm(envvars, 4),
            data=datlist[[i]]$traindat, 
            family="binomial")
  varimp <- biomod2::variables_importance(
    gamStart, 
    data=traindatlist[[i]]$traindat)$mat
  varimp <- varimp[varimp[,1]!=0,]
  df2 <- rbind(df2, data.frame(varimp=varimp, term=names(varimp), i=paste("samp", i)))
}
```

```{r}
p <- ggplot(df2, aes(term, varimp)) +
  geom_boxplot() + coord_flip() +
  ggtitle("Variable importance across data sets")
p
```


# Models across many seeds

Now rerun the analysis with many different random splits of the data and see how variable the results are.

```{r eval=FALSE}
n <- length(traindatlist)
allmodels <- list()
for(mn in 1:n){
  cat(mn," ") #progress

  traindat <- traindatlist[[i]]$traindat

modellist <- list()

glmVIF <- stats::glm(formula =pa ~ ., 
                family = binomial(link = "logit"),
                data=traindat[,c("pa", vifvars)])
modellist[["glmVIF"]] <- list(model=glmVIF, name="glmVIF", desc="GLM - all VIF variables", terms=vifvars)

glmStep <- MASS::stepAIC(object=glmVIF, 
                     scope=pa ~ .,
                     data = traindat.sub,
                     direction = "both", 
                     trace = -1, #quiet
                     k = 2, 
                     control=glm.control(maxit=500))
modellist[["glmStep"]] <- list(model=glmStep, name="glmStep", desc="GLM - variables selected w step AIC", terms=attr(glmStep$terms, "term.labels"))

glmLC <- stats::glm(formula =pa ~ ., 
                family = binomial(link = "logit"),
                data=traindat[,c("pa",lcvars)])
modellist[["glmLC"]] <- list(model=glmLC, name="glmLC", desc="GLM - LC variables", terms=lcvars)

glmTopo <- stats::glm(formula =pa ~ ., 
                family = binomial(link = "logit"),
                data=traindat[,c("pa",topovars)])
modellist[["glmTopo"]] <- list(model=glmTopo, name="glmTopo", desc="GLM - topo variables", terms=topovars)

glmEnv <- stats::glm(formula =pa ~ ., 
                family = binomial(link = "logit"),
                data=traindat[,c("pa",envvars)])
modellist[["glmEnv"]] <- list(model=glmEnv, name="glmEnv", desc="GLM - environmental variables", terms=envvars)

gamEnvMin <- gam::gam(formula = gamfm(minEnvVars, 4),
            data=traindat, 
            family="binomial")
modellist[["gamEnvMin"]] <- list(model=gamEnvMin, name="gamEnvMin", desc="GAM - Minimal", terms=minEnvVars)

gamTopo2 <- gam::gam(formula = gamfm(topovars, 2),
            data=traindat, 
            family="binomial")
gamTopo4 <- gam::gam(formula = gamfm(topovars, 4),
            data=traindat, 
            family="binomial")
mod <- paste0("gamTopo", c(2,4))
desc <- paste0("GAM - Topo variables df=", c(2,4))
for(i in 1:2){
modellist[[mod[i]]] <- list(model=get(mod[i]), name=mod[i], desc=desc[i], terms=topovars)
}

gamLC2 <- gam::gam(formula = gamfm(lcvars,2),
            data=traindat, 
            family="binomial")
gamLC4 <- gam::gam(formula = gamfm(lcvars,4),
            data=traindat, 
            family="binomial")

mod <- paste0("gamLC", c(2,4))
desc <- paste0("GAM - Tree Cover variables df=", c(2,4))
for(i in 1:2){
modellist[[mod[i]]] <- list(model=get(mod[i]), name=mod[i], desc=desc[i], terms=topovars)
}

gamEnv2 <- gam::gam(formula = gamfm(envvars, 2),
            data=traindat, 
            family="binomial")
gamEnv4 <- gam::gam(formula = gamfm(envvars, 4),
            data=traindat, 
            family="binomial")
mod <- paste0("gamEnv", c(2,4))
desc <- paste0("GAM - Environmental variables df=", c(2,4))
for(i in 1:2){
modellist[[mod[i]]] <- list(model=get(mod[i]), name=mod[i], desc=desc[i], terms=topovars)
}
for(i in names(modellist)){
  pm <- predict(allVars,
              modellist[[i]]$model,
              type="response")
bindex <- ecospat::ecospat.boyce(pm, cbind(pres_test$lon, pres_test$lat), PEplot=FALSE)
modellist[[i]]$bindex <- bindex
modellist[[i]]$AIC <- AIC(modellist[[i]]$model)
modellist[[i]]$AUC <- dismo::evaluate(pres_test, backg_test, model=modellist[[i]]$model)@auc
modellist[[i]]$Spearman.cor <- modellist[[i]]$bindex$Spearman.cor
}
allmodels[[mn]] <- modellist
}
save(allmodels, file="allmodels.RData")
df2 <- data.frame()
for(i in 1:length(allmodels)){
  minAIC <- min(unlist(lapply(allmodels[[i]], function(x){x$AIC})))
  for(j in names(allmodels[[i]])){
    mod <- allmodels[[i]][[j]]
  df2 <- rbind(df2, data.frame(n=i,model=j,delAIC=mod$AIC-minAIC, Spearman.cor=mod$Spearman.cor, AUC=mod$AUC, AIC=mod$AIC))
  }
}
df2$n <- as.factor(df$n)
df.allmodels <- df2
save(df.allmodels, allmodels, file="allmodels.RData")
save(df.allmodels, file="dfallmodels.RData")
```

```{r}
library(tidyr)
load("dfallmodels.RData")
df.long <- pivot_longer(df.allmodels, c(-n, -model), names_to = "term", values_to = "value")
p <- ggplot(df.long, aes(model, value))
p + geom_boxplot() + coord_flip() + facet_wrap(~term, scale="free_x")
```


